Requirements Specification

### 2025-05-14 nbs/development/requirements.qmd
docs(development): propagate model selection updates to requirements
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: "Requirements Specification"
format:
  html:
    mermaid-format: js
    mermaid:
      theme: dark
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
---

@@

PyroVelocity is a probabilistic framework for RNA velocity analysis in single-cell genomics. It provides a Bayesian approach to modeling RNA dynamics, enabling robust uncertainty quantification and parameter inference from single-cell RNA sequencing data. The system implements a modular architecture with interchangeable components for dynamics models, prior distributions, likelihood models, observation models, and inference guides.

@@

1. Directly modeling raw spliced and unspliced read counts
2. Providing uncertainty quantification for velocity estimates
3. Synchronizing cell time estimation across genes
4. Enabling multivariate modeling of gene expression
5. Offering diagnostic analyses and visualizations of model uncertainty
6. Integrating with existing single-cell analysis ecosystems

The system is currently transitioning from a legacy implementation to a modular PyTorch/Pyro implementation, with a future migration to a JAX/NumPyro implementation for improved performance.

@@

- **Single-cell RNA-seq Data**: Source data containing spliced and unspliced counts
- **AnnData Objects**: Primary data structure for storing and manipulating single-cell data
- **Scanpy/scVelo**: Ecosystem tools for preprocessing and visualization
- **Pyro/PyTorch**: Core probabilistic programming and tensor computation frameworks
- **JAX/NumPyro**: Alternative backend for improved performance (future)

```{mermaid}
graph TB
    User["Data Scientist / Researcher"] -->|uses| Library["PyroVelocity"]
    Library -->|depends on| PyTorch["PyTorch"]
    Library -->|depends on| Pyro["Pyro"]
    Library -->|processes| AnnData["AnnData Objects"]
    Library -->|produces| Velocity["RNA Velocity"]
    Library -->|integrates with| Scanpy["Scanpy"]
    Library -->|integrates with| Scvelo["scVelo"]

    class User user;
    class Library library;
    class PyTorch,Pyro,Scanpy,Scvelo external;
    class AnnData,Velocity data;
```

@@

| Capability ID | Capability Name | Description | Priority |
|--------------|-----------------|-------------|----------|
| CAP-01 | Probabilistic RNA Velocity | Bayesian modeling of RNA velocity with uncertainty quantification | High |
| CAP-02 | Parameter Inference | Inference of biologically meaningful parameters from scRNA-seq data | High |
| CAP-03 | Uncertainty Visualization | Visualization of uncertainty in velocity estimates and trajectories | High |
| CAP-04 | AnnData Integration | Seamless integration with AnnData objects and the scanpy ecosystem | High |
| CAP-05 | Modular Architecture | Flexible, component-based architecture with interchangeable parts | High |
| CAP-06 | JAX/NumPyro Backend | High-performance implementation using JAX/NumPyro | Medium |
| CAP-07 | Workflow Integration | Integration with workflow systems for reproducible analysis | Medium |

@@

- **Data Preprocessing**: Scientists prepare single-cell data for velocity analysis
- **Model Configuration**: Scientists configure model components and parameters
- **Model Training**: Scientists train the model using variational inference
- **Posterior Analysis**: Scientists analyze posterior distributions of parameters
- **Velocity Computation**: Scientists compute RNA velocity with uncertainty
- **Result Visualization**: Scientists visualize velocity and uncertainty
- **Integration with Workflows**: Scientists integrate velocity analysis into larger pipelines

```{mermaid}
graph TD
    A[Data Preprocessing] --> B[Model Configuration]
    B --> C[Model Training]
    C --> D[Posterior Analysis]
    D --> E[Velocity Computation]
    E --> F[Result Visualization]
    F --> G[Integration with Workflows]

    class A,B,C,D,E,F,G scenario;
```

@@

| Scenario ID | Scenario Name | Primary Actor | Description | Pre-conditions | Post-conditions |
|-------------|---------------|---------------|-------------|----------------|-----------------|
| US-01 | Data Preprocessing | A-01, A-02 | Prepare single-cell data for velocity analysis | Raw count data available | AnnData object with spliced/unspliced counts |
| US-02 | Model Configuration | A-01, A-04 | Configure model components and parameters | Preprocessed data available | Configured model ready for training |
| US-03 | Model Training | A-01, A-02 | Train the model using variational inference | Configured model and data available | Trained model with optimized parameters |
| US-04 | Posterior Analysis | A-01, A-04 | Analyze posterior distributions of parameters | Trained model available | Parameter distributions and diagnostics |
| US-05 | Velocity Computation | A-01, A-02 | Compute RNA velocity with uncertainty | Posterior samples available | Velocity estimates with uncertainty |
| US-06 | Result Visualization | A-01, A-03 | Visualize velocity and uncertainty | Velocity estimates available | Visualizations for interpretation |
| US-07 | Integration with Workflows | A-02 | Integrate velocity analysis into larger pipelines | Trained model and analysis code available | Integrated workflow |

@@

**Usage Scenario:** US-03 (Model Training)

@@

1. User loads preprocessed AnnData object with spliced/unspliced counts
2. User creates a PyroVelocity model with standard components
3. System initializes model parameters and guide
4. User calls the train method with the AnnData object
5. System performs stochastic variational inference
6. System monitors convergence and reports progress
7. System returns the trained model
8. User saves the trained model for later use

@@

* At step 2, if user wants custom components, they configure specific components
* At step 4, if user wants custom training parameters, they provide additional arguments

@@

* At step 1, if data is not properly formatted, system provides error message
* At step 5, if training fails to converge, system provides diagnostic information

@@

**Workflow Name:** RNA Velocity Analysis Workflow

**Description:** Complete workflow for RNA velocity analysis from raw data to interpretation

**Participants:** A-01, A-02, A-03

@@

1. Data Preprocessing: US-01
2. Model Configuration: US-02
3. Model Training: US-03
4. Posterior Analysis: US-04
5. Velocity Computation: US-05
6. Result Visualization: US-06

@@

* For large datasets, users may subsample genes or cells before model training
* For complex biological systems, users may use nonlinear dynamics models
* For integration with other analyses, users may export velocity to other tools

@@

- Optimized guide parameters
- Training diagnostics
- ELBO values

@@

- Time Complexity: O(num_epochs * num_cells * num_genes)
- Space Complexity: O(num_cells * num_genes)
- Convergence: Depends on model complexity and data characteristics

@@

- Pyro documentation
- Hoffman et al. (2013)

@@

@@

| Parameter | Type | Description | Constraints |
|-----------|------|-------------|------------|
| adata | AnnData | Annotated data object | Must contain spliced/unspliced layers |
| max_epochs | int | Maximum number of training epochs | > 0 |
| batch_size | Optional[int] | Batch size for training | None or > 0 |
| learning_rate | float | Learning rate for optimizer | > 0 |
| use_gpu | bool | Whether to use GPU acceleration | True/False |

@@

| Parameter | Type | Description |
|-----------|------|-------------|
| model | PyroVelocityModel | Trained model |
| loss_history | List[float] | Training loss history |

@@

| Error Code | Description | Recovery Action |
|------------|-------------|----------------|
| ValueError | Invalid AnnData format | Preprocess data correctly |
| RuntimeError | Training divergence | Adjust learning rate or model parameters |
| MemoryError | Out of memory | Reduce batch size or use data subsampling |

@@

- Parallelization: Data parallelism across cells and genes
- Distribution: Single-node execution with optional GPU acceleration
- Scheduling: Batch-based processing for large datasets

@@

- Compute: Multi-core CPU or CUDA-compatible GPU
- Memory: 16-64GB RAM depending on dataset size
- Storage: 10-100GB for datasets and results
- Network: Minimal (data loading only)

@@

- Scales linearly with number of cells and genes
- GPU acceleration provides significant speedup for large datasets
- JAX/NumPyro implementation offers improved performance over PyTorch/Pyro

@@

| Algorithm ID | Parameter | Description | Target Value | Measurement Method |
|--------------|-----------|-------------|--------------|-------------------|
| ALG-01 | Execution Time | Time to compute velocity for 10k cells, 2k genes | < 1 minute on GPU | Wall clock time |
| ALG-02 | Training Time | Time to train model on 10k cells, 2k genes | < 30 minutes on GPU | Wall clock time |
| ALG-02 | Memory Usage | Peak memory usage during training | < 16GB | Process monitoring |
| ALG-02 | Convergence Rate | Number of epochs to reach convergence | < 1000 epochs | ELBO monitoring |

@@

| Data Object ID | Data Object Name | Description | Related Scientific Data Object | Persistence |
|----------------|------------------|-------------|-------------------------------|------------|
| DO-01 | AnnData | Annotated data matrix for single-cell data | AnnData | Yes |
| DO-02 | CountMatrix | Matrix of RNA counts | CountMatrix | Yes |
| DO-03 | ModelParameters | Parameters of the RNA velocity model | ModelParameters | Yes |
| DO-04 | PosteriorSamples | Samples from parameter posterior distributions | PosteriorSamples | Yes |
| DO-05 | VelocityEstimates | RNA velocity vectors for each cell | VelocityEstimates | Yes |
| DO-06 | ModelState | Immutable state container for model components | N/A | Yes |
| DO-07 | ComponentConfig | Configuration for model components | N/A | Yes |

@@

@@

| Relationship ID | From Object | To Object | Type | Cardinality | Description |
|-----------------|-------------|-----------|------|-------------|-------------|
| REL-01 | AnnData | CountMatrix | Composition | 1:N | AnnData contains multiple count matrices (X, layers) |
| REL-02 | PyroVelocityModel | ModelParameters | Association | 1:1 | Model uses parameters for computation |
| REL-03 | PyroVelocityModel | PosteriorSamples | Association | 1:1 | Model generates posterior samples |
| REL-04 | PosteriorSamples | VelocityEstimates | Derivation | 1:1 | Velocity is computed from posterior samples |
| REL-05 | AnnData | VelocityEstimates | Association | 1:1 | Velocity estimates are stored in AnnData |
| REL-06 | PyroVelocityModel | ModelState | Composition | 1:1 | Model contains state |
| REL-07 | ModelState | ComponentConfig | Composition | 1:N | State contains component configurations |

```{mermaid}
classDiagram
    AnnData *-- CountMatrix : contains
    PyroVelocityModel o-- ModelParameters : uses
    PyroVelocityModel --> PosteriorSamples : generates
    PosteriorSamples --> VelocityEstimates : computes
    AnnData o-- VelocityEstimates : stores
    PyroVelocityModel *-- ModelState : contains
    ModelState *-- ComponentConfig : contains

>>>>

---
title: "Requirements Specification"
format:
  html:
    mermaid-format: js
    mermaid:
      theme: dark
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
bibliography: ../references.bib
csl: ../bibstyle.csl
---

@@

PyroVelocity is a probabilistic framework for RNA velocity analysis in single-cell genomics. It provides a Bayesian approach to modeling RNA dynamics, enabling robust uncertainty quantification, parameter inference, and model comparison from single-cell RNA sequencing data. The system implements a modular architecture with interchangeable components for dynamics models, prior distributions, likelihood models, observation models, and inference guides.

@@

1. Directly modeling raw spliced and unspliced read counts
2. Providing uncertainty quantification for velocity estimates
3. Synchronizing cell time estimation across genes
4. Enabling multivariate modeling of gene expression
5. Offering diagnostic analyses and visualizations of model uncertainty
6. Supporting model validation, comparison, and selection
7. Integrating with existing single-cell analysis ecosystems

The system is currently transitioning from a legacy implementation to a modular PyTorch/Pyro implementation, with a future migration to a JAX/NumPyro implementation for improved performance and support for multiple inference algorithms.

@@

- **Single-cell RNA-seq Data**: Source data containing spliced and unspliced counts
- **AnnData Objects**: Primary data structure for storing and manipulating single-cell data
- **Scanpy/scVelo**: Ecosystem tools for preprocessing and visualization
- **Pyro/PyTorch**: Core probabilistic programming and tensor computation frameworks
- **JAX/NumPyro**: Alternative backend for improved performance and inference algorithms
- **torchode/diffrax**: Differential equation solvers for dynamical simulation
- **beartype/jaxtyping**: Runtime type checking and array shape annotations
- **Flyte/hydra-zen**: Workflow orchestration and configuration management

```{mermaid}
graph TB
    User["Data Scientist / Researcher"] -->|uses| Library["PyroVelocity"]
    Library -->|depends on| PyTorch["PyTorch"]
    Library -->|depends on| Pyro["Pyro"]
    Library -->|depends on| JAX["JAX"]
    Library -->|depends on| NumPyro["NumPyro"]
    Library -->|depends on| Torchode["torchode"]
    Library -->|depends on| Diffrax["diffrax"]
    Library -->|depends on| Beartype["beartype"]
    Library -->|depends on| Jaxtyping["jaxtyping"]
    Library -->|depends on| Flyte["Flyte"]
    Library -->|depends on| HydraZen["hydra-zen"]
    Library -->|processes| AnnData["AnnData Objects"]
    Library -->|produces| Velocity["RNA Velocity"]
    Library -->|produces| ModelComparison["Model Comparison"]
    Library -->|integrates with| Scanpy["Scanpy"]
    Library -->|integrates with| Scvelo["scVelo"]

    class User user;
    class Library library;
    class PyTorch,Pyro,JAX,NumPyro,Torchode,Diffrax,Beartype,Jaxtyping,Flyte,HydraZen,Scanpy,Scvelo external;
    class AnnData,Velocity,ModelComparison data;
```

@@

| Capability ID | Capability Name | Description | Priority |
|--------------|-----------------|-------------|----------|
| CAP-01 | Probabilistic RNA Velocity | Bayesian modeling of RNA velocity with uncertainty quantification | High |
| CAP-02 | Parameter Inference | Inference of biologically meaningful parameters from scRNA-seq data using multiple inference algorithms | High |
| CAP-03 | Uncertainty Visualization | Visualization of uncertainty in velocity estimates, parameters, and trajectories | High |
| CAP-04 | AnnData Integration | Seamless integration with AnnData objects and the scanpy ecosystem | High |
| CAP-05 | Modular Architecture | Flexible, component-based architecture with interchangeable parts for systematic model comparison | High |
| CAP-06 | Parameter Recovery Validation | Validation of model's ability to recover known parameters from synthetic data to ensure empirically identifiable models | High |
| CAP-07 | Model Comparison | Systematic comparison of different model architectures, implementations, and inference algorithms | High |
| CAP-08 | JAX/NumPyro Backend | High-performance implementation using JAX/NumPyro with support for MCMC inference | Medium |
| CAP-09 | Workflow Integration | Integration with workflow systems for reproducible analysis and model validation | Medium |
| CAP-10 | Dynamical Simulation | Support for numerical integration of ODEs for non-linear and stochastic dynamics | Medium |

@@

- **Data Preprocessing**: Scientists prepare single-cell data for velocity analysis
- **Model Configuration**: Scientists configure model components and parameters
- **Parameter Recovery Validation**: Scientists validate model's ability to recover known parameters from synthetic data
- **Model Training**: Scientists train the model using variational inference or MCMC
- **Posterior Analysis**: Scientists analyze posterior distributions of parameters
- **Velocity Computation**: Scientists compute RNA velocity with uncertainty
- **Result Visualization**: Scientists visualize velocity, parameters, and associated uncertainty
- **Model Comparison**: Scientists systematically compare different model architectures, implementations, and inference algorithms
- **Model Selection**: Scientists select the most appropriate model based on validation metrics
- **Integration with Workflows**: Scientists integrate velocity analysis and model validation into larger pipelines

```{mermaid}
graph TD
    A[Data Preprocessing] --> B[Model Configuration]
    B --> V[Parameter Recovery Validation]
    V --> C[Model Training]
    C --> D[Posterior Analysis]
    D --> E[Velocity Computation]
    E --> F[Result Visualization]
    F --> M[Model Comparison]
    M --> S[Model Selection]
    S --> G[Integration with Workflows]

    class A,B,V,C,D,E,F,M,S,G scenario;
```

@@

| Scenario ID | Scenario Name | Primary Actor | Description | Pre-conditions | Post-conditions |
|-------------|---------------|---------------|-------------|----------------|-----------------|
| US-01 | Data Preprocessing | A-01, A-02 | Prepare single-cell data for velocity analysis | Raw count data available | AnnData object with spliced/unspliced counts |
| US-02 | Model Configuration | A-01, A-04 | Configure model components and parameters | Preprocessed data available | Configured model ready for training |
| US-03 | Parameter Recovery Validation | A-01, A-04 | Validate model's ability to recover known parameters from synthetic data | Model specification and prior distributions available | Validation metrics and visualizations demonstrating parameter recovery performance |
| US-04 | Model Training | A-01, A-02 | Train the model using variational inference or MCMC | Configured model and data available | Trained model with optimized parameters |
| US-05 | Posterior Analysis | A-01, A-04 | Analyze posterior distributions of parameters | Trained model available | Parameter distributions and diagnostics |
| US-06 | Velocity Computation | A-01, A-02 | Compute RNA velocity with uncertainty | Posterior samples available | Velocity estimates with uncertainty |
| US-07 | Result Visualization | A-01, A-03 | Visualize velocity, parameters, and associated uncertainty | Velocity estimates and parameter distributions available | Visualizations for interpretation |
| US-08 | Model Comparison | A-01, A-04 | Systematically compare different model architectures, implementations, and inference algorithms | Multiple trained models available | Comparison metrics and visualizations |
| US-09 | Model Selection | A-01, A-04 | Select the most appropriate model based on validation metrics | Model comparison results available | Selected model for downstream analysis |
| US-10 | Integration with Workflows | A-02 | Integrate velocity analysis and model validation into larger pipelines | Trained model, selected model, and analysis code available | Integrated workflow |

@@

**Usage Scenario:** US-03 (Parameter Recovery Validation)

**Description:** Validating a PyroVelocity model's ability to recover known parameters from synthetic data to ensure empirically identifiable models

**Triggering Event:** User initiates parameter recovery validation

1. User specifies model configuration, prior distributions, and inference algorithm (SVI or MCMC)
2. System generates known parameter sets from prior distributions
3. System creates synthetic data using these parameters and the model's generative process
4. System performs inference on synthetic data to estimate parameters
5. System compares estimated parameters to original known values
6. System analyzes recovery accuracy, precision, and coverage
7. System calculates credible intervals and coverage metrics
8. System generates validation metrics and visualizations
9. User evaluates parameter recovery performance
10. System stores validation results for model comparison

* At step 1, if user wants to test specific parameter values, they provide custom parameter sets
* At step 1, if user wants to compare inference algorithms, they specify multiple algorithms
* At step 3, if user wants to test robustness, they specify noise levels to add to synthetic data
* At step 4, if user wants to test multiple sample sizes, they specify a range of sample sizes

* At step 4, if inference fails to converge, system provides diagnostic information
* At step 5, if parameters are not recoverable, system identifies problematic parameters
* At step 7, if credible intervals have poor coverage, system suggests model refinement

**Scenario ID:** FS-02

**Usage Scenario:** US-04 (Model Training)

@@

1. User loads preprocessed AnnData object with spliced/unspliced counts
2. User creates a PyroVelocity model with standard components
3. User selects inference algorithm (SVI or MCMC)
4. System initializes model parameters and inference components
5. User calls the train method with the AnnData object
6. If using SVI:
   a. System performs stochastic variational inference
   b. System monitors ELBO convergence and reports progress
7. If using MCMC:
   a. System performs Markov Chain Monte Carlo sampling
   b. System monitors chain diagnostics and reports progress
8. System returns the trained model with posterior samples
9. User saves the trained model for later use

@@

* At step 2, if user wants custom components, they configure specific components
* At step 3, if user wants to compare inference algorithms, they specify multiple algorithms
* At step 5, if user wants custom training parameters, they provide additional arguments

@@

* At step 1, if data is not properly formatted, system provides error message
* At step 6a, if SVI fails to converge, system provides diagnostic information
* At step 7a, if MCMC chains fail to mix, system suggests parameter adjustments

@@

**Workflow Name:** RNA Velocity Analysis Workflow with Model Validation and Selection

**Description:** Complete workflow for RNA velocity analysis from raw data to interpretation, including parameter recovery validation, model comparison, and model selection

**Participants:** A-01, A-02, A-03, A-04

@@

1. Data Preprocessing: US-01
2. Model Configuration: US-02
3. Parameter Recovery Validation: US-03
4. Model Training: US-04
5. Posterior Analysis: US-05
6. Velocity Computation: US-06
7. Result Visualization: US-07
8. Model Comparison: US-08
9. Model Selection: US-09
10. Integration with Workflows: US-10 (optional)

@@

* For large datasets, users may subsample genes or cells before model training
* For complex biological systems, users may use nonlinear dynamics models with numerical integration
* For integration with other analyses, users may export velocity to other tools
* For comprehensive validation, users may test multiple parameter sets, sample sizes, and noise levels
* For model comparison, users may compare different model architectures, implementations, and inference algorithms
* For inference algorithm comparison, users may compare SVI and MCMC approaches

**Workflow ID:** WF-02

**Workflow Name:** Parameter Recovery Validation and Model Comparison Workflow

**Description:** Focused workflow for validating model's ability to recover known parameters from synthetic data and comparing different model architectures and inference algorithms

**Participants:** A-01, A-04

1. Model Configuration: US-02
2. Parameter Recovery Validation: US-03
3. Result Visualization: US-07
4. Model Comparison: US-08
5. Model Selection: US-09

* For comprehensive validation, users may test multiple parameter sets, sample sizes, and noise levels
* For robustness testing, users may vary sample sizes, noise levels, and data generation processes
* For model development, users may iterate on model design based on validation results
* For inference algorithm comparison, users may compare SVI and MCMC approaches
* For empirical identifiability assessment, users may analyze parameter recovery across different model architectures

@@

- Optimized guide parameters
- Training diagnostics
- ELBO values
- Posterior samples

@@

- Time Complexity: O(num_epochs * num_cells * num_genes)
- Space Complexity: O(num_cells * num_genes)
- Convergence: Depends on model complexity and data characteristics
- Parallelization: Embarrassingly parallel across batches

@@

- Pyro documentation [@Bingham2018-id]
- Hoffman et al. (2013)

**Algorithm ID:** ALG-04

**Algorithm Name:** Markov Chain Monte Carlo Inference

**Description:** Sampling-based Bayesian inference for RNA velocity parameters

```
p(z|x) ∝ p(x|z)p(z)
```

Where:
- p(z|x) is the posterior distribution of latent variables z given observed data x
- p(x|z) is the likelihood of observed data given latent variables
- p(z) is the prior distribution of latent variables

- Model: Probabilistic model p(x, z)
- Data: Observed data x
- Num_samples: Number of posterior samples
- Num_chains: Number of parallel chains
- Num_warmup: Number of warmup steps

- Posterior samples
- Chain diagnostics (R-hat, effective sample size)
- Posterior predictive checks

- Time Complexity: O(num_samples * num_chains * num_cells * num_genes)
- Space Complexity: O(num_samples * num_chains * num_parameters)
- Convergence: Assessed via chain diagnostics
- Parallelization: Embarrassingly parallel across chains

- NumPyro documentation [@Phan2019-numpyro]
- Hoffman and Gelman (2014) [No U-Turn Sampler]

@@

**Description:** Interface for parameter recovery validation

| Parameter | Type | Description | Constraints |
|-----------|------|-------------|------------|
| model_config | Dict[str, Any] | Model configuration | Must specify model components |
| prior_config | Dict[str, Any] | Prior distribution configuration | Must specify prior distributions for all parameters |
| inference_algorithm | str | Inference algorithm to use (SVI or MCMC) | "SVI" or "MCMC" |
| num_parameter_sets | int | Number of parameter sets to test | > 0 |
| num_samples | int | Number of synthetic datasets per parameter set | > 0 |
| sample_sizes | List[int] | List of sample sizes to test | All values > 0 |
| noise_levels | List[float] | List of noise levels to test | All values >= 0 |
| use_gpu | bool | Whether to use GPU acceleration | True/False |

| Parameter | Type | Description |
|-----------|------|-------------|
| validation_results | Dict[str, Any] | Validation metrics and results |
| parameter_recovery_plots | Dict[str, Figure] | Visualizations of parameter recovery |
| recovery_summary | Dict[str, float] | Summary statistics of recovery performance |
| credible_interval_coverage | Dict[str, float] | Coverage statistics for credible intervals |

| Error Code | Description | Recovery Action |
|------------|-------------|----------------|
| ValueError | Invalid model or prior configuration | Correct configuration |
| RuntimeError | Inference failure | Adjust inference parameters or model complexity |
| MemoryError | Out of memory | Reduce number of parameter sets or sample sizes |
| ConvergenceError | Inference algorithm fails to converge | Adjust inference parameters or try alternative algorithm |

**Interface ID:** CI-03

**Algorithm:** ALG-03, ALG-04

@@

| Parameter | Type | Description | Constraints |
|-----------|------|-------------|------------|
| adata | AnnData | Annotated data object | Must contain spliced/unspliced layers |
| inference_algorithm | str | Inference algorithm to use (SVI or MCMC) | "SVI" or "MCMC" |
| max_epochs | int | Maximum number of training epochs (SVI) or samples (MCMC) | > 0 |
| batch_size | Optional[int] | Batch size for training (SVI only) | None or > 0 |
| learning_rate | float | Learning rate for optimizer (SVI only) | > 0 |
| num_chains | int | Number of MCMC chains (MCMC only) | > 0 |
| num_warmup | int | Number of warmup steps (MCMC only) | > 0 |
| use_gpu | bool | Whether to use GPU acceleration | True/False |

@@

| Parameter | Type | Description |
|-----------|------|-------------|
| model | PyroVelocityModel | Trained model |
| loss_history | List[float] | Training loss history (SVI only) |
| posterior_samples | Dict[str, Tensor] | Posterior samples |
| diagnostics | Dict[str, Any] | Training or sampling diagnostics |

@@

| Error Code | Description | Recovery Action |
|------------|-------------|----------------|
| ValueError | Invalid AnnData format | Preprocess data correctly |
| RuntimeError | Training divergence or sampling failure | Adjust parameters or try alternative algorithm |
| MemoryError | Out of memory | Reduce batch size, number of chains, or use data subsampling |
| ConvergenceError | Inference algorithm fails to converge | Adjust inference parameters or try alternative algorithm |

@@

- Parallelization: Data parallelism across cells and genes
- Distribution: Single-node execution with optional GPU acceleration
- Scheduling: Batch-based processing for large datasets
- Inference: Multiple inference algorithms (SVI and MCMC)

@@

- Compute: Multi-core CPU or CUDA-compatible GPU
- Memory: 16-64GB RAM depending on dataset size and inference algorithm
- Storage: 10-100GB for datasets, results, and posterior samples
- Network: Minimal (data loading only)

@@

- Scales linearly with number of cells and genes
- GPU acceleration provides significant speedup for large datasets
- JAX/NumPyro implementation offers improved performance over PyTorch/Pyro
- MCMC requires more computational resources than SVI but provides more accurate uncertainty quantification
- Parallel chains in MCMC can leverage multi-GPU systems

@@

| Algorithm ID | Parameter | Description | Target Value | Measurement Method |
|--------------|-----------|-------------|--------------|-------------------|
| ALG-01 | Execution Time | Time to compute velocity for 10k cells, 2k genes | < 1 minute on GPU | Wall clock time |
| ALG-02 | Training Time | Time to train model on 10k cells, 2k genes | < 30 minutes on GPU | Wall clock time |
| ALG-02 | Memory Usage | Peak memory usage during training | < 16GB | Process monitoring |
| ALG-03 | Convergence Rate (SVI) | Number of epochs to reach convergence | < 1000 epochs | ELBO monitoring |
| ALG-04 | Sampling Time (MCMC) | Time to generate 1000 posterior samples with 4 chains | < 2 hours on GPU | Wall clock time |
| ALG-04 | Memory Usage (MCMC) | Peak memory usage during MCMC sampling | < 32GB | Process monitoring |
| ALG-04 | Chain Convergence | R-hat statistic for all parameters | < 1.1 | MCMC diagnostics |
| ALG-04 | Effective Sample Size | Minimum effective sample size across parameters | > 100 | MCMC diagnostics |

@@

| Data Object ID | Data Object Name | Description | Related Scientific Data Object | Persistence |
|----------------|------------------|-------------|-------------------------------|------------|
| DO-01 | AnnData | Annotated data matrix for single-cell data | AnnData | Yes |
| DO-02 | CountMatrix | Matrix of RNA counts | CountMatrix | Yes |
| DO-03 | ModelParameters | Parameters of the RNA velocity model | ModelParameters | Yes |
| DO-04 | PosteriorSamples | Samples from parameter posterior distributions | PosteriorSamples | Yes |
| DO-05 | VelocityEstimates | RNA velocity vectors for each cell | VelocityEstimates | Yes |
| DO-06 | ModelState | Immutable state container for model components | N/A | Yes |
| DO-07 | ComponentConfig | Configuration for model components | N/A | Yes |
| DO-08 | SyntheticData | Synthetic data generated with known parameters | N/A | Yes |
| DO-09 | ParameterRecoveryResults | Results of parameter recovery validation | N/A | Yes |
| DO-10 | ModelComparisonResults | Results of comparing different model architectures and inference algorithms | N/A | Yes |
| DO-11 | MCMCDiagnostics | Diagnostics for MCMC sampling | N/A | Yes |
| DO-12 | CredibleIntervals | Credible intervals for parameter estimates | N/A | Yes |
| DO-13 | ValidationMetrics | Metrics for validating model performance | N/A | Yes |

@@

**Data Object:** DO-08 (SyntheticData)

| Attribute ID | Attribute Name | Description | Type | Units | Constraints | Default Value |
|--------------|----------------|-------------|------|-------|-------------|---------------|
| ATTR-15 | u_obs | Observed unspliced counts | Tensor | Counts | Non-negative | None |
| ATTR-16 | s_obs | Observed spliced counts | Tensor | Counts | Non-negative | None |
| ATTR-17 | true_parameters | True parameter values used to generate data | Dict[str, Tensor] | Various | Non-negative | None |
| ATTR-18 | sample_size | Number of cells in the synthetic dataset | int | Count | > 0 | None |
| ATTR-19 | noise_level | Noise level used in data generation | float | Dimensionless | >= 0 | 0.0 |
| ATTR-20 | seed | Random seed used for reproducibility | int | Dimensionless | Any | None |

**Data Object:** DO-09 (ParameterRecoveryResults)

| Attribute ID | Attribute Name | Description | Type | Units | Constraints | Default Value |
|--------------|----------------|-------------|------|-------|-------------|---------------|
| ATTR-21 | true_parameters | True parameter values | Dict[str, Tensor] | Various | Non-negative | None |
| ATTR-22 | estimated_parameters | Estimated parameter values | Dict[str, Tensor] | Various | Non-negative | None |
| ATTR-23 | credible_intervals | Credible intervals for estimated parameters | Dict[str, Tuple[Tensor, Tensor]] | Various | Lower < Upper | None |
| ATTR-24 | recovery_metrics | Metrics quantifying parameter recovery performance | Dict[str, Dict[str, float]] | Various | None | None |
| ATTR-25 | coverage_metrics | Metrics quantifying credible interval coverage | Dict[str, float] | Percentage | [0, 100] | None |
| ATTR-26 | bias_metrics | Metrics quantifying systematic bias in estimates | Dict[str, float] | Various | None | None |
| ATTR-27 | sample_size | Number of cells used in recovery | int | Count | > 0 | None |
| ATTR-28 | noise_level | Noise level used in data generation | float | Dimensionless | >= 0 | 0.0 |
| ATTR-29 | inference_algorithm | Inference algorithm used | str | N/A | "SVI" or "MCMC" | "SVI" |
| ATTR-30 | model_architecture | Model architecture used | str | N/A | Valid model name | "standard" |
| ATTR-31 | empirical_identifiability | Assessment of parameter identifiability | Dict[str, bool] | N/A | True/False | None |

**Data Object:** DO-10 (ModelComparisonResults)

| Attribute ID | Attribute Name | Description | Type | Units | Constraints | Default Value |
|--------------|----------------|-------------|------|-------|-------------|---------------|
| ATTR-32 | model_configs | Configurations of compared models | List[Dict[str, Any]] | N/A | Non-empty | None |
| ATTR-33 | inference_algorithms | Inference algorithms used | List[str] | N/A | "SVI" or "MCMC" | None |
| ATTR-34 | performance_metrics | Performance metrics for each model | Dict[str, Dict[str, float]] | Various | None | None |
| ATTR-35 | parameter_recovery_metrics | Parameter recovery metrics for each model | Dict[str, Dict[str, float]] | Various | None | None |
| ATTR-36 | uncertainty_calibration | Uncertainty calibration metrics for each model | Dict[str, Dict[str, float]] | Various | None | None |
| ATTR-37 | computational_metrics | Computational performance metrics | Dict[str, Dict[str, float]] | Various | None | None |
| ATTR-38 | selected_model | Selected model based on comparison | str | N/A | Valid model name | None |
| ATTR-39 | selection_criteria | Criteria used for model selection | Dict[str, float] | Various | None | None |

**Data Object:** DO-11 (MCMCDiagnostics)

| Attribute ID | Attribute Name | Description | Type | Units | Constraints | Default Value |
|--------------|----------------|-------------|------|-------|-------------|---------------|
| ATTR-40 | r_hat | R-hat convergence statistic | Dict[str, float] | Dimensionless | > 0 | None |
| ATTR-41 | effective_sample_size | Effective sample size | Dict[str, float] | Count | > 0 | None |
| ATTR-42 | chain_statistics | Statistics for each chain | Dict[str, Dict[str, Any]] | Various | None | None |
| ATTR-43 | acceptance_rate | MCMC acceptance rate | float | Percentage | [0, 100] | None |
| ATTR-44 | divergences | Number of divergent transitions | int | Count | >= 0 | None |
| ATTR-45 | energy_statistics | Energy statistics for HMC | Dict[str, float] | Various | None | None |

@@

| Relationship ID | From Object | To Object | Type | Cardinality | Description |
|-----------------|-------------|-----------|------|-------------|-------------|
| REL-01 | AnnData | CountMatrix | Composition | 1:N | AnnData contains multiple count matrices (X, layers) |
| REL-02 | PyroVelocityModel | ModelParameters | Association | 1:1 | Model uses parameters for computation |
| REL-03 | PyroVelocityModel | PosteriorSamples | Association | 1:1 | Model generates posterior samples |
| REL-04 | PosteriorSamples | VelocityEstimates | Derivation | 1:1 | Velocity is computed from posterior samples |
| REL-05 | AnnData | VelocityEstimates | Association | 1:1 | Velocity estimates are stored in AnnData |
| REL-06 | PyroVelocityModel | ModelState | Composition | 1:1 | Model contains state |
| REL-07 | ModelState | ComponentConfig | Composition | 1:N | State contains component configurations |
| REL-08 | ModelParameters | SyntheticData | Generation | 1:N | Parameters are used to generate synthetic data |
| REL-09 | SyntheticData | ParameterRecoveryResults | Association | 1:1 | Synthetic data is used for parameter recovery validation |
| REL-10 | ParameterRecoveryResults | ModelParameters | Validation | N:1 | Recovery results validate parameter inference |
| REL-11 | PosteriorSamples | MCMCDiagnostics | Association | 1:1 | MCMC diagnostics are generated from posterior samples |
| REL-12 | PosteriorSamples | CredibleIntervals | Derivation | 1:1 | Credible intervals are computed from posterior samples |
| REL-13 | ParameterRecoveryResults | ModelComparisonResults | Association | N:1 | Parameter recovery results are used for model comparison |
| REL-14 | ModelComparisonResults | ValidationMetrics | Composition | 1:1 | Model comparison results contain validation metrics |
| REL-15 | PyroVelocityModel | ModelComparisonResults | Association | N:1 | Multiple models are compared in model comparison |
